---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
      collapse = TRUE,
      comment = "#>",
      fig.path = "man/figures/README-",
      out.width = "100%",
      dpi = 300
)
```

# ggcube

<!-- badges: start -->
<!-- badges: end -->

The ggcube library is an extension of ggplot2 for 3D plotting.

This package is under development, and a stable version has not yet been released. 


## Example

To use ggcube, you use standard ggplot2 code along with a `projection()` and one or more `proj_*()` layer functions.

Let's make a 3D scatterplot of the `mpg` dataset. We'll use `proj_data(geom = "point")` to add our main data, and `proj_margin()` to add some marginal scatter plots as well. ggplot2 is flexible and powerful, but a downside to using it for 3D plots is that in addition to adding our data, we also have to explicitly build up the reference layers like the panel background, gridlines, and labels. 

```{r example, warning = FALSE, message = FALSE}
library(ggcube)

# get data
d <- select(mpg, displ, hwy, cty, cyl)

# define projection
prj <- projection(pitch = -20, roll = 160, 
                  persp = TRUE, dist = 2, data = d)

# build plot
d %>%
      ggplot(aes(displ, hwy, z = cty, color = factor(cyl))) +
      proj_panel(prj = prj) +
      proj_gridlines(prj = prj) +
      proj_margin(prj = prj, geom = "point", 
                  color = "gray40", size = 1, alpha = .5) +
      proj_data(prj = prj, geom = "point", size = 3) +
      proj_label(prj = prj, axis = "x", title = "displacement",
                 edge = c("ymin", "zmin")) +
      proj_label(prj = prj, axis = "y", title = "highway MPG",
                 edge = c("zmax", "xmax")) +
      proj_label(prj = prj, axis = "z", title = "city MPG",
                 edge = c("xmin", "ymax")) +
      coord_fixed() +
      theme_void()
```


As another example, let's plot the `mountain` data set that comes with the package, using the
`proj_ridgeline()` function to visualize a 3D surface:

```{r example2, warning = FALSE, message = FALSE}
prj <- projection(pitch = 75, yaw = 55,
                  persp = T, dist = 1,
                  data = mountain)

mountain %>%
      ggplot(aes(x, y, z = z)) +
      proj_ridgeline(prj = prj, geom = "polygon", piece = "y",
                     fill = "black", color = "white", linewidth = .15) +
      theme_void()
```



## Installation

You can install the development version of ggcube from [GitHub](https://github.com/matthewkling/ggcube) with:

``` r
devtools::install_github("matthewkling/ggcube")
```
