---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
      collapse = TRUE,
      comment = "#>",
      fig.path = "man/figures/README-",
      out.width = "100%",
      dpi = 300
)
```

# ggcube

<!-- badges: start -->
<!-- badges: end -->

The `ggcube` R library is an extension of ggplot2 for 3D plotting. This package is under development, and a stable version has not yet been released. 

The package's core function is `coord_3d()`, which provides a 3D coordinate system for use in combination with `aes(x, y, z)`. The library also offers a number of 3D stats and geoms to plot objects like surfaces and volumes. 


## Example

Let's start with a super basic 3D scatter plot of the `mpg` dataset. Compared to a standard 2D ggplot, we simply expand the aesthetic mapping to include our third variable, and then add `coord_3d()`:

```{r example, warning = FALSE, message = FALSE}
library(ggplot2)
library(ggcube)

ggplot(mpg, aes(displ, hwy, cty, color = drv)) +  # Enhanced aes() from ggcube
      geom_point() + # From ggplot2
      coord_3d() # From ggcube
```

But the library also provides its own `geom_point_3d` function that includes features like scaling point size by perceived distance, and adding reference points projected as 2D scatterplots on the panel faces:

```{r scatter, warning = FALSE, message = FALSE}
ggplot(mpg, aes(displ, hwy, cty, color = drv)) +
      geom_point_3d(
            ref_points = TRUE,
            ref_faces = c("xmax", "ymin", "zmin"),
            ref_point_alpha = .25) + 
      coord_3d()
```

A key 3D stat is `stat_surface()`, which can be used to plot grid-like data that has a z-value for every x-y combination, such as function predictions or terrain layers. The example below uses this stat to visualize a wave function; in addition to projecting the surface, it uses some 3D `lighting()` to highlight the surface shape. It also hints at how you can adjust the rotation and aspect ratios of your plot via arguments to `coord_3d()`, and apply ggplot2 themes to the plot.

```{r surface, warning = FALSE, message = FALSE}
# data
library(dplyr)
library(tidyr)
d <- expand_grid(x = seq(-10, 10, .25),
                 y = seq(-10, 10, .25)) %>%
      mutate(z = cos(sqrt(x^2 + y^2)) * 3)

# plot
ggplot(d, aes(x, y, z)) +
      coord_3d(pitch = 0, roll = 130, yaw = 30, scales = "fixed") +
      stat_surface(fill = "orangered", color = "orangered", linewidth = .1,
                   light = lighting(blend = "both", blend_mode = "hsl")) +
      theme_dark()
```

You can also create an arbitrary third axis along which to array a set of data. Here's an example using a map of the US:

```{r layers, warning = FALSE, message = FALSE}
# data
country <- map_data("usa") %>% mutate(level = "country")
state <- map_data("state") %>% mutate(level = "state")
county <- map_data("county") %>% mutate(level = "county")
d <- bind_rows(country, state, county) %>%
      mutate(level = factor(level, levels = c("county", "state", "country"))) %>%
      mutate(group = paste0(level, "_", group)) %>%
      arrange(group, order)

# plot
d %>%
      ggplot(aes(long, lat, level, fill = level, group = group)) +
      geom_polygon_3d(color = "black", linewidth = .05) +
      coord_3d(scales = "fixed", ratio = c(1, 1.3, 5),
               pitch = 0, roll = -60, yaw = 30) +
      theme(legend.position = "none")
```





## Installation

You can install the development version of ggcube from [GitHub](https://github.com/matthewkling/ggcube) with:

``` r
devtools::install_github("matthewkling/ggcube")
```
