---
title: "Controlling the 3D view"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Controlling the 3D view}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 150,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(ggcube)
library(patchwork)
```

`coord_3d()` is the core of every ggcube plot. It controls how 3D data is rotated,
projected, framed, and decorated. This article walks through each of its parameters
in detail.

We'll use a common base plot throughout:

```{r base}
p <- ggplot() +
  geom_function_3d(
    aes(fill = after_stat(z), color = after_stat(z)),
    fun = function(x, y) sin(x) * cos(y),
    xlim = c(-pi, pi), ylim = c(-pi, pi),
    n = 50, light = light("direct", contrast = .7)) +
  scale_fill_viridis_c() +
  scale_color_viridis_c() +
  theme_light() +
  theme(legend.position = "none",
        axis.text = element_blank())
```


## Rotation

Three angles control how the plot is oriented: `pitch`, `roll`, and `yaw` rotate the
plot around the x-, y-, and z-axes, respectively. All are specified in degrees. When
they're all zero, the plot resembles a standard 2D plot (+x = right, +y = up, +z = 
toward viewer). The defaults (`pitch = 0`, `roll = -60`, `yaw = -30`) give an overhead
corner view that can work well for many plots.

Here are a variety of rotations showing the default, the zero-rotation case, the
effect of setting each rotation angle individually while the others are held at zero,
and an example combining them to achieve an arbitrary rotation (note this uses the
`patchwork` library to produce a mutlti-panel plot):
 
```{r rotation}
(p + coord_3d() + ggtitle("Default")) +
  (p + coord_3d(pitch = 0, roll = 0, yaw = 0) + ggtitle("All angles 0")) +
  (p + coord_3d(pitch = 20, roll = 40, yaw = 60) + ggtitle("Arbitrary combination")) +
  (p + coord_3d(pitch = 30, roll = 0, yaw = 0) + ggtitle("pitch = 30")) +
  (p + coord_3d(pitch = 0, roll = 30, yaw = 0) + ggtitle("roll = 30")) +
  (p + coord_3d(pitch = 0, roll = 0, yaw = 30) + ggtitle("yaw = 30")) +
  plot_layout(nrow = 2)
```

 
 
## Perspective and distance
 
By default, `coord_3d()` uses perspective projection, where objects farther from the
viewer appear smaller. The `dist` parameter controls the camera distance from the center
of the data cube --- larger values produce less distortion. Set `persp = FALSE` to switch
to orthographic projection, where parallel lines in 3D remain parallel in the 2D
rendering. Orthographic projection is equivalent to viewing from infinite distance and can be useful
for technical or schematic plots.
 
```{r perspective, fig.height = 3}
(p + coord_3d(yaw = 45, dist = 1.1) + ggtitle("dist = 1.1")) +
  (p + coord_3d(yaw = 45, dist = 2) + ggtitle("dist = 2")) +
  (p + coord_3d(yaw = 45, persp = FALSE) + ggtitle("persp = FALSE")) +
  plot_layout(ncol = 3)
```
 
 
 
## Aspect ratios and scale matching
 
The `scales` parameter controls how data ranges map to visual cube dimensions. With
`"free"` (the default), each axis is stretched independently to fill its dimension of the
cube, maximizing the visual range for each variable. With `"fixed"`, the visual
proportions match the data proportions, analogous to `coord_fixed()` in 2D:
 
```{r scales, fig.height = 3}
(p + coord_3d(scales = "free") + ggtitle('scales = "free"')) +
  (p + coord_3d(scales = "fixed") + ggtitle('scales = "fixed"')) +
  plot_layout(ncol = 2) & theme_light()
```
 
The `ratio` parameter sets custom relative lengths for the three axes. With `"free"`
scales, ratios are applied to the standardized cube coordinates. With `"fixed"` scales,
they're applied to the data coordinates:
 
```{r ratio, fig.height = 3.5}
(p + coord_3d(scales = "free", ratio = c(1, 3, 1)) +
    ggtitle('free, ratio = c(1, 3, 1)')) +
  (p + coord_3d(scales = "fixed", ratio = c(1, 3, 1)) +
      ggtitle('fixed, ratio = c(1, 3, 1)')) +
  plot_layout(ncol = 2) & theme_light()
```


## Panel selection
 
The `panels` argument controls which faces of the cube are drawn. The six faces are
named `"xmin"`, `"xmax"`, `"ymin"`, `"ymax"`, `"zmin"`, and `"zmax"`. Depending on the
rotation, some faces lie behind the data ("background" panels) and some lie in
front ("foreground" panels). By default, only background panels are drawn. When drawn,
foreground panels default to 20% opacity so they don't obscure the data; their styling
is controlled via theme elements (see the theming section below).
 
You can request specific faces by name, show all faces, or hide them entirely:
 
```{r panels}
(p + coord_3d(panels = "background") + ggtitle('"background" (default)')) +
  (p + coord_3d(panels = c("xmin", "xmax", "zmax")) +
      ggtitle('c("xmin", "xmax", "zmax")')) +
  (p + coord_3d(panels = "all") + ggtitle('"all"')) +
  (p + coord_3d(panels = "none") + ggtitle('"none"')) +
  plot_layout(ncol = 2)
```

## Zooming, clipping, and expansion
 
`zoom` adjusts the overall framing without changing the rotation or projection. Values less than 1 zoom
out (more whitespace), while values greater than 1 zoom in (tighter framing).

The `clip` parameter controls whether graphical elements can be drawn outside the panel area (ggplot2's
rectangular canvas, not the projected 3D panel). The default `"off"` allows this, which is generally 
recommended for 3D plots where elements may extend beyond the projected panel area.
 
```{r zoom, fig.height = 3}
(p + coord_3d(zoom = 0.7) + ggtitle("zoom = 0.7")) +
  (p + coord_3d(zoom = 1) + ggtitle("zoom = 1\n(default)")) +
  (p + coord_3d(zoom = 1.5, clip = "on") + ggtitle('zoom = 1.5\nclip = "on"')) +
  plot_layout(ncol = 3)
```
The `expand` parameter works like the `expand` argument in standard ggplot2 scales ---
when `TRUE` (the default), axis ranges are padded slightly beyond the data range. You can
fine-tune expansion per axis using the standard scale functions.
 
```{r expand, fig.height = 4}
(p + coord_3d(expand = TRUE) + ggtitle("expand = TRUE (default)")) +
  (p + coord_3d(expand = FALSE) + ggtitle("expand = FALSE")) +
  plot_layout(ncol = 2)
```


 
## Axis label placement
 
The `xlabels`, `ylabels`, and `zlabels` parameters control where axis tick labels and
titles are placed. Each axis potentially has multiple possible edges where labels could 
appear (depending on which faces are visible). The default `"auto"` uses a heuristic that
prioritizes edges on the periphery of the plot for readability.
 
For manual control, pass a length-2 character vector naming two adjacent faces --- the
shared edge between them is where the labels go. The first face determines the alignment
plane:
 
```{r labels, fig.height = 4}
(p + coord_3d() + ggtitle("auto (default)")) +
  (p + coord_3d(xlabels = c("ymax", "zmax"),
                zlabels = c("xmax", "ymin")) +
      ggtitle("manual placement")) +
  plot_layout(ncol = 2) & theme_light()
```
 
The `rotate_labels` parameter controls whether labels automatically rotate to align with
the projected axis direction. Setting it to `FALSE` uses the angle specified in theme
settings instead:
 
```{r labels_rotate, fig.height = 4}
(p + coord_3d(rotate_labels = TRUE) + ggtitle("rotate_labels = TRUE")) +
  (p + coord_3d(rotate_labels = FALSE) + ggtitle("rotate_labels = FALSE")) +
  plot_layout(ncol = 2) & theme_light()
```
 
The `title_position` parameter affects axis titles that fall on internal (non-peripheral)
edges. `"auto"` places them at the near end of the axis outside the plot area, while
`"center"` centers them along the edge.

 
 
## Theming the 3D cube
 
Standard ggplot2 themes work with ggcube. Adding a complete theme behaves as expected:
 
```{r themes, fig.height = 3}
(p + coord_3d() + theme_dark() + ggtitle("theme_dark()")) +
  (p + coord_3d() + theme_minimal() + ggtitle("theme_minimal()")) +
  plot_layout(ncol = 2)
```

ggcube also adds several theme elements for 3D-specific styling. One important category is
foreground panel elements, which control cube faces rendered in front of the data.
These inherit from their background counterparts, so styling `panel.background` affects
both background and foreground faces, while `panel.foreground` targets only the front
faces. ggcube's `element_rect()` extends ggplot2's version with an `alpha` parameter for
transparency. This is particularly useful for foreground panels --- the default
`panel.foreground` uses `alpha = 0.2` so front faces don't obscure the data:
 
```{r theme_foreground}
p +
  coord_3d(panels = "all") +
  theme_gray() +
  theme(panel.border = element_rect(color = "black"),
        panel.foreground = element_rect(alpha = .3),
        panel.grid.foreground = element_line(color = "gray", linewidth = .25))
```

For axis labels, ggcube adds `axis.text.z` and `axis.title.z` elements that inherit from
`axis.text` and `axis.title` respectively. 

The `margin` argument to `element_text()` controls padding between titles, labels, and 
cube edges. Because ggcube's current system for text spacing is fairly crude, margins need
to be manually adjusted more often than in base ggplot2 figures.
 
```{r theme_axes}
p +
  coord_3d(panels = "all") +
  theme(panel.background = element_rect(color = "black"),
        panel.border = element_rect(color = "black"),
        panel.foreground = element_rect(alpha = .3),
        panel.grid.foreground = element_line(color = "gray", linewidth = .25),
        axis.text = element_text(color = "darkblue", size = 12),
        axis.text.z = element_text(color = "darkred"),
        axis.title = element_text(margin = margin(t = 30)),
        axis.title.x = element_text(color = "magenta"))
```
 
The full set of 3D-specific theme elements:
 
- **`panel.foreground`**: Fill and border for foreground cube faces (inherits from
  `panel.background`)
- **`panel.border.foreground`**: Border for foreground faces (inherits from
  `panel.border`)
- **`panel.grid.foreground`**: Grid lines on foreground faces (inherits from
  `panel.grid`)
- **`panel.grid.major.foreground`**: Major grid lines on foreground faces (inherits from
  `panel.grid.foreground`)
- **`axis.text.z`**: Z-axis tick labels (inherits from `axis.text`)
- **`axis.title.z`**: Z-axis title (inherits from `axis.title`)
