---
title: "Lighting and shading"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lighting and shading}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 150
)
```

```{r setup, message = FALSE}
library(ggcube)
library(patchwork)
```

Lighting modifies the fill and/or color of polygon faces based on their orientation
relative to a light source, giving surfaces and volumes a sense of depth and shape.
It applies to any polygon-based 3D layer --- surfaces, hulls, bars, voxels, and polygon
text.

We'll use a common base plot throughout:

```{r base}
p <- ggplot(sphere_points, aes(x, y, z)) +
  geom_hull_3d(fill = "#9e2602", color = "#5e1600") +
  coord_3d(scales = "fixed")
p
```



## Lighting targets

### Layers

Lighting can be set at two levels:

- **Coord-level**: `coord_3d(light = light(...))` applies to all layers in the plot.
- **Layer-level**: `geom_*_3d(light = light(...))` overrides the coord-level setting for
  that layer.

Use `light = "none"` to disable lighting entirely, or `light = NULL` in a layer to
inherit the coord-level setting (the default layer behavior):

```{r where}
p + geom_hull_3d(aes(x = x + 2.5), fill = "#9e2602", color = "#5e1600",
                 light = "none")
```

### Aesthetic targets

The `fill` and `color` parameters control which aesthetics get modified by lighting. 
By default both are `TRUE`:

```{r targets}
(p + ggtitle("fill + color (default)")) +
  (p + coord_3d(light = light(fill = TRUE, color = FALSE)) +
      ggtitle("fill only"))
```

Lighting works with both solid and aesthetically-mapped colors:

```{r mapped_colors}
ggplot(sphere_points, aes(x, y, z)) +
  geom_hull_3d(aes(fill = x, color = x)) +
  scale_fill_viridis_c() +
  scale_color_viridis_c() +
  guides(fill = guide_colorbar_3d()) +
  coord_3d(light = light("direct"))
```

## Light qualities

### Lighting methods

The `method` parameter controls the lighting model:

- **`"diffuse"`** (default): Soft, atmospheric lighting. Only surfaces pointing directly
  away from the light source are fully dark.
- **`"direct"`**: Hard lighting. All surfaces angled beyond 90 degrees from the light
  are fully dark.
- **`"rgb"`**: Maps surface orientation to RGB color space, replacing the base colors
  entirely.
 

```{r methods}
(p + coord_3d(light = light(method = "diffuse")) + ggtitle('"diffuse" (default)')) +
  (p + coord_3d(light = light(method = "direct")) + ggtitle('"direct"')) +
  (p + coord_3d(light = light(method = "rgb")) + ggtitle('"rgb"')) +
  plot_layout(ncol = 3)
```




### Color modes

The `mode` parameter controls how lighting modifies colors. This applies to the
`"diffuse"` and `"direct"` methods (not `"rgb"`). There are two color modes:

- **`"hsv"`** (default): Modifies the value component of HSV color. Highlights preserve
  hue; shadows fade to black.
- **`"hsl"`**: Modifies the lightness component of HSL color. Highlights fade to white;
  shadows fade to black.

The two modes give identical results for grayscale base colors.

```{r modes}
(p + coord_3d(light = light(mode = "hsv")) +
    ggtitle('"hsv" (default)')) +
  (p + coord_3d(light = light(mode = "hsl")) +
      ggtitle('"hsl"'))
```


### Contrast strength

The `contrast` parameter scales the intensity of lighting effects. Values less than 1
give subtler effects; values greater than 1 give more dramatic ones:

```{r contrast}
(p + coord_3d(light = light(mode = "hsl", contrast = 0.5)) +
    ggtitle("contrast = 0.5")) +
  (p + coord_3d(light = light(mode = "hsl", contrast = 1)) +
      ggtitle("contrast = 1 (default)")) +
  (p + coord_3d(light = light(mode = "hsl", contrast = 1.5)) +
      ggtitle("contrast = 1.5")) +
  plot_layout(ncol = 3)
```



## Light sources

### Direction and anchoring

The `direction` parameter is a length-3 vector specifying where light comes from; for example, 
`direction = c(0, -1, 0)` lights the plot from the negative y direction. The `anchor` parameter 
controls whether this direction is relative to the data axes or the camera:

- **`anchor = "scene"`** (default): Direction is relative to the data axes and rotates
  with the plot. `direction = c(1, 0, 0)` always lights surfaces facing the "xmax" face.
- **`anchor = "camera"`**: Direction is fixed relative to the viewing pane.
  `direction = c(1, 0, 0)` always lights surfaces facing the right side of the plot,
  regardless of rotation.

```{r direction}
(p + coord_3d(light = light("direct", direction = c(0, -1, 0))) +
    ggtitle('anchor = "scene"')) +
  (p + coord_3d(light = light("direct", direction = c(0, -1, 0), anchor = "camera")) +
      ggtitle('anchor = "camera"'))
```


### Positional lighting

In addition to directional lighting (parallel rays from a direction), ggcube also supports
positional lighting from a point source. Use `position` instead of `direction` to specify
a light source location in data coordinates. Setting `distance_falloff = TRUE` applies 
inverse-square intensity falloff from the light source position.

```{r positional}
ggplot(mountain, aes(x, y, z)) +
  stat_surface_3d(fill = "red", color = "red") +
  coord_3d(
    light = light(position = c(.5, .7, 95), distance_falloff = TRUE,
                  mode = "hsl", contrast = .9),
    ratio = c(1.5, 2, 1))
```


## Backface lighting

A backface is the side of a polygon facing the underside of a surface or the inside of a
volume. By default, backfaces receive inverted lighting relative to frontfaces 
(`backface_scale = -1`), creating strong visual contrast between the two sides. 
Setting `backface_scale = 1` lights backfaces the same as frontfaces. Modifying 
`backface_offset` to uniformly darkens or brightens all backfaces.


```{r backface}
pb <- ggplot() +
  geom_function_3d(fun = function(x, y) x^2 + y^2,
                   xlim = c(-3, 3), ylim = c(-3, 3),
                   fill = "steelblue", color = "steelblue")

(pb + coord_3d(pitch = 0, roll = -70, yaw = 0,
               light = light(mode = "hsl")) +
    ggtitle("default:\nbackface_scale = -1\nbackface_offset = 0")) +
  (pb + coord_3d(pitch = 0, roll = -70, yaw = 0,
                 light = light(backface_scale = 1, mode = "hsl")) +
      ggtitle("\nbackface_scale = 1\nbackface_offset = 0")) +
  (pb + coord_3d(pitch = 0, roll = -70, yaw = 0,
              light = light(backface_scale = 1, mode = "hsl",
                            backface_offset = -.5)) +
      ggtitle("\nbackface_scale = 1\nbackface_offset = -0.5")) +
  plot_layout(nrow = 1)
```



## 3D guides

When lighting is active, standard color guides don't reflect the range of shaded colors
visible in the plot. ggcube provides `guide_colorbar_3d()` and `guide_legend_3d()` to
create guides that show the full shading gradient:

```{r guide_colorbar}
ggplot(mountain, aes(x, y, z, fill = z)) +
  stat_surface_3d(light = light(mode = "hsl", direction = c(1, 0, 0))) +
  guides(fill = guide_colorbar_3d()) +
  scale_fill_gradientn(colors = c("tomato", "dodgerblue")) +
  coord_3d(ratio = c(2, 3, 1.5))
```

```{r guide_legend}
ggplot(mountain, aes(x, y, z, fill = x > .5, group = 1)) +
  stat_surface_3d(light = light(mode = "hsl", direction = c(1, 0, 0))) +
  guides(fill = guide_legend_3d()) +
  coord_3d(ratio = c(2, 3, 1.5))
```

Both guides accept `shade_range` to control the range of shading shown (a length-2
vector in the range -1 to 1), and `reverse_shade` to flip the gradient direction.

Note that when fill and color map to the same variable, ggplot2 creates a shared scale.
In this case, apply the 3D guide via `guides(fill = guide_colorbar_3d())`, not via the
color guide.
