---
title: "Surfaces"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 150,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, message = FALSE}
library(ggcube)
library(patchwork)
```

Surfaces are ggcube's richest feature area. This article gives an overview of the geoms and 
stats that work together to render continuous surfaces from various data sources. "Surface" here
means a data object with a single z value for each (x, y) position, like a heightmap or 
terrain model. This vignette doesn't cover hulls (`geom_hull_3d()`) or columnar surfaces 
(`geom_col_3d()`). 

Surfaces can be rendered from several data sources and in several visual styles, using a 
system of interchangeable geoms and stats. ggcube currently includes three surface geoms:

- `geom_surface_3d()`: tessellated mesh, rendered as a solid surface or a wireframe
- `geom_contour_3d()`: stacks of elevation contour polygons
- `geom_ridgeline_3d()`: arrays of horizontal cross-section polygons

Each of these geoms can render data produced by one of four surface stats:

- `stat_surface_3d()`: user-supplied irregular or gridded data
- `stat_function_3d()`: mathematical functions
- `stat_smooth_3d()`: statistical models fit to user-supplied data
- `stat_density_3d()`: 2D kernel density surface


## Data sources

Surface data comes in two forms:

- **Regular grids** have a z value at every combination of x and y positions --- like a
raster or DEM. Grid data can be supplied directly by the user, or generated internally by
a stat.
- **Irregular points** are scattered (x, y, z) observations with no grid structure. These
are triangulated via Delaunay tessellation to produce a surface mesh.

`stat_surface_3d()` auto-detects which type of data it receives and handles both. All 
three surface geoms work with regular grid data, but only `geom_surface_3d()` supports 
irregular points (since ridgelines and contours require an underlying grid structure).


## Surface geoms

Three geoms render surface data in different visual styles.

### Tessellated mesh

`geom_surface_3d()` is the primary surface geom. You can use it to create a solid
surface as shown here, or set `fill = NA` to produce a wireframe. It tessellates 
data into polygon tiles and supports both regular grids and irregular point data:

```{r geom_surface}
p <- ggplot(mountain, aes(x, y, z, fill = z, color = z)) +
  coord_3d(ratio = c(1, 1.5, .75)) +
  scale_fill_viridis_c() +
  scale_color_viridis_c() +
  theme_light()

p + geom_surface_3d(light = light(direction = c(1, 0, .5),
                                mode = "hsv", contrast = 1.5),
                  linewidth = .2) +
  guides(fill = guide_colorbar_3d())
```

### Contours

`geom_contour_3d()` creates filled contour bands stacked at their respective elevations,
like a topographic layer-cake. The `bins` or `breaks` parameters control the contour
levels:

```{r geom_contour}
p + geom_contour_3d(bins = 12, color = "white", light = "none")
```


### Ridgelines

`geom_ridgeline_3d()` slices a surface into cross-sections, producing a ridgeline
view. The `direction` parameter controls the slicing axis:

```{r geom_ridgeline}
p + stat_surface_3d(geom = "ridgeline_3d", direction = "y",
                    fill = "black", color = "white",
                    light = "none", linewidth = .1)
```


### Mixing stats and geoms

Each of these geoms can be paired with any of the surface stats described below. For example,
`stat_function_3d(geom = "contour_3d")` renders a mathematical function as contour
bands, and `stat_function_3d(geom = "ridgeline_3d")` renders it as ridgelines.


## Surface stats

Four stats produce surface data. Each generates a regular grid of (x, y, z) points
that can be rendered by any of the three surface geoms.

### User-supplied data

`stat_surface_3d()` is the default stat for `geom_surface_3d()`. It works with regular grids,
and with irregular point data for which it performs Delaunay triangulation as shown here:

```{r stat_surface}
set.seed(42)
pts <- data.frame(x = runif(200, -2, 2), y = runif(200, -2, 2))
pts$z <- with(pts, sin(x) * cos(y))

ggplot(pts, aes(x, y, z = z, fill = z)) +
  stat_surface_3d(sort_method = "pairwise") +
  scale_fill_viridis_c() +
  coord_3d(light = "none")
```

### Mathematical functions

`stat_function_3d()` evaluates a function f(x, y) over a grid:

```{r stat_function}
ggplot(mapping = aes(color = after_stat(x))) +
  geom_function_3d(fun = function(x, y) sin(x) * cos(y),
                   xlim = c(-pi, pi), ylim = c(-pi, pi),
                   fill = NA) + # disable fill to make wireframe
  coord_3d(light = "none")
```

### Statistical models

`stat_smooth_3d()` (or `geom_smooth_3d()`) fits a model to scattered (x, y, z) data
and renders the fitted surface. It has options for different model types, for overlaying
data point and residual lines, and for adding confidence interval surfaces as shown here:

```{r stat_smooth}
ggplot(mtcars, aes(qsec, disp, mpg, fill = after_stat(level))) +
      geom_smooth_3d(domain = "chull", se = TRUE, color = "black") +
      scale_fill_manual(values = c("red", "darkorchid4", "steelblue")) +
      coord_3d(yaw = 150) + theme_light()
```

### Kernel density

`stat_density_3d()` (or `geom_density_3d()`) computes a 2D kernel density estimate.
It has options for modifying bandwidth, resolution:

```{r stat_density}
ggplot(faithful, aes(eruptions, waiting)) +
  geom_density_3d(min_ndensity = .01) +
  guides(fill = guide_colorbar_3d()) +
  coord_3d() +
  scale_fill_viridis_c()
```


## Working with surface meshes

The following options apply to `geom_surface_3d()` and the tessellated mesh it produces.

### Grid types

The `grid` parameter controls tile geometry. The default `"rectangle"` produces
rectangular tiles. `"right1"` and `"right2"` split each rectangle into right triangles
along opposite diagonals, and `"equilateral"` produces an equilateral triangular lattice.
Triangulated grids can prevent lighting artifacts on sharply curving surfaces:

```{r grid_types, fig.width = 8, fig.height = 8}
d <- dplyr::mutate(tidyr::expand_grid(x = -10:10, y = -10:10),
                   z = sqrt(x^2 + y^2) / 1.5,
                   z = cos(z) - z)
p <- ggplot(d, aes(x, y, z)) +
  coord_3d(light = light(mode = "hsl", direction = c(1, 0, 0)))

(p + geom_surface_3d(fill = "white", color = "darkblue",
                      linewidth = .2) +
    ggtitle('"rectangle" (default)')) +
  (p + geom_surface_3d(fill = "white", color = "darkblue",
                        linewidth = .2, grid = "right1") +
      ggtitle('"right1"')) +
  (p + geom_surface_3d(fill = "white", color = "darkblue",
                        linewidth = .2, grid = "right2") +
      ggtitle('"right2"')) +
  (p + geom_surface_3d(fill = "white", color = "darkblue",
                        linewidth = .2, grid = "equilateral") +
      ggtitle('"equilateral"')) +
  plot_layout(ncol = 2)
```

### Computed variables

Surface stats compute gradient information at each grid point, available via
`after_stat()`: partial derivatives (`dzdx`, `dzdy`), gradient magnitude (`slope`), and
direction of steepest ascent (`aspect`):

```{r computed_vars}
p + geom_surface_3d(aes(fill = after_stat(slope)), grid = "right2") +
  scale_fill_viridis_c() +
  guides(fill = guide_colorbar_3d())
```

### Lighting

Surfaces often benefit from ggcube's lighting system, which modifies polygon face
colors based on their orientation relative to a light source. Note that since 
contours and ridgelines have uniform polygon orientation, they typically do
not benefit from lighting. See `vignette("lighting")` for details.

```{r lighting, fig.height = 3.5}
p <- ggplot(mountain, aes(x, y, z)) +
  coord_3d(ratio = c(1, 1.5, .75)) +
  scale_fill_viridis_c() + scale_color_viridis_c() +
  theme(legend.position = "none")

(p + geom_surface_3d(aes(fill = z, color = z), light = "none") +
    ggtitle('light = "none"')) +
  (p + geom_surface_3d(aes(fill = z, color = z),
                        light = light(direction = c(1, 0, 0))) +
      ggtitle("diffuse (default)")) +
  plot_layout(ncol = 2)
```

