---
title: "Animated and interactive rotation"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dpi = 150,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(ggcube)
```

Because ggcube renders a 3D scene from a fixed camera angle, a single plot only
ever shows one viewpoint. Two functions let you show more than one:
`animate_3d()` moves the camera through a sequence of angles and plays them
back as a GIF or movie, while `orbit_3d()` renders the same kind of sequence
(or a 2D grid of rotation angles) into a HTML "orbit" widget you can drag 
to rotate. 

Animations and orbits can be displayed in RStudio's viewer, exported and viewed in a 
web browser, or knit into RMarkdown documents like Reveal.js presentations or
pkgdown articles. They can also be added to Shiny apps using 

Both functions simply render the plot many times at different rotations, so they 
preserve ggcube's exact look --- every frame is a real ggcube render, with the 
same lighting, depth sorting, and color scales. 

The two functions share a rotation vocabulary. They operate on a pre-existing
ggcube plot object, and take `pitch`, `roll`, and `yaw` arguments that define
rotation ranges to move through.

We'll use a common base plot throughout this article:

```{r base}
p <- ggplot(mountain, aes(x, y, z, fill = z, color = z)) +
  geom_surface_3d(linewidth = .25) +
  coord_3d(light = light(anchor = "camera", direction = c(1, 1, 0)),
           ratio = c(1.5, 2, 1), zoom = 1.25) +
  scale_fill_viridis_c() +
  scale_color_viridis_c() +
  guides(fill = guide_colorbar_3d()) +
  theme_void()
```


## Animations

`animate_3d()` takes a plot and a set of keyframe angles, interpolates between
them, and renders the frames into an animation. Each of `pitch`, `roll`, and
`yaw` is either held at a single value or given as a vector of keyframes to move
through; the camera is linearly interpolated between successive keyframes.

The most common case is a turntable spin --- a full `yaw` rotation, holding the
other angles fixed. `nframes` sets the number of rendered frames and `fps` the
playback speed:

```{r animate-spin}
animate_3d(p, yaw = c(0, 360), nframes = 60, fps = 10, width = 700)
```

Keyframes can also trace a more complex path. Here the view moves through
several angles on two axes at once, tilting on the `roll` axis while it spins 
in `yaw`:

```{r animate-path}
animate_3d(p,
           yaw = c(0, 720),
           roll = c(-90, 0, -90),
           nframes = 120, fps = 10, width = 700)
```

By default the animation is written to a temporary file and displayed, but you can
use `file =` to save it.

See `?animate_3d` for the full set of playback options, including frame timing, 
pauses, rewind, alternative renderers, and parallel rendering.


## Interactive orbits

`orbit_3d()` also renders a set of plots at different rotation angles, but it instead 
packages the frames into an interactive HTML widget that you can drag to rotate. Clicking 
and dragging with your cursor, or using your keyboard arrow keys (after clicking or 
tab-selecting the plot), changes the viewing angle. 

Rotation is specified the same way as in animation, except that here each angle 
is given as a range rather than a keyframe path. Specifying the range in ascending vs.
descending order switches the control direction; note that for `yaw` specifically, 
specifying the larger angle first (e.g. `c(360, 0)`) is generally preferred so that 
rotation direction matches drag direction.

Full-turn ranges (e.g. `yaw = c(0, 360)`) wrap around seamlessly, so you can keep 
dragging in one direction forever, while partial ranges stop at their endpoints. 
Use `start =` to choose the initial angle. Smoothness is governed by the frame count `n`, 
with the trade-off that more frames take longer to render and produce larger files.

Specifying a single range gives a 1D orbit or turntable effect --- drag left and right 
to spin it:

```{r orbit-1d}
orbit_3d(p, yaw = c(360, 0), start = 300,
            n = 36, width = 700)
```

For a 2D orbit, you specify two ranges; horizontal drag turns the plot on one axis 
and vertical drag on the other, letting you both spin and tilt the plot. 
The two numbers in `n` set the resolution of each axis, and their product determines the
number of rendered frames, so keep them modest:

```{r orbit-2d}
orbit_3d(p, yaw = c(360, 0), roll = c(-90, 0), start = c(yaw = 300),
            n = c(24, 12), width = 700)
```

You can include orbit widgets in Shiny apps via `orbit3dOutput()` and `renderorbit3d()`; 
see the documentation for those functions to learn more.


## Computational considerations

Pre-rendering dozens or hundreds of frames can take some time, and can generate
large files. If needed, you can speed up rendering using parallelization, via 
the `cores` parameter. You can also reduce the number of frames with the `n` 
parameter, and change the size of the resulting images with `width`/`height`.


